### 2020-08-18 manuscript/manuscript.rmd
update manuscript based on Alex's comments from google doc

---
bibliography: manuscript.bib
csl: nature.csl
output:
  word_document: default
  pdf_document: 
    latex_engine: xelatex
  html_document: default
  github_document: default
---

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 4). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

Other approaches have been taken to addressing base compositional heterogeneity across taxa. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. However, Our own analyses (Supplementary Information section S3, Supplementary Figs. 5-7) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition. Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence. Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many? Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.

@@

#### Sensitive analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were used as the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), with the exception of different numbers of categories were used.

To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Posson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above. The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.

@@

We thank the Yale Center for Research Computing for use of the research computing infrastructure, specificaly the Farnam HPC cluster.

@@

**Extended Data Fig. 3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sampled isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis. The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses. (C,D) MDS plots for the amino acid frequencies. Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies. A single MDS analysis was done for all analyses, so that the projections are the same in each plot. For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories. This adds a considerable number of parameters for describing a similar overall space.

@@

**Extended Data Fig. 4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

@@

Note that there was, in fact, an exception to strong support. An analysis of the 40 ribosomal proteins in the matrix recovered Ctenophora-sister with only 69% support. This study did not include *Trichoplax*.

@@

Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore phylogenetic relationship of early diverging animals by added 9 new species. Th data matrix were assembled based on The phylogenetic analysis using Poisson+CAT model strongly supported Porifera-sister, and ctenophores was sister to cnidarians to form the "coelenterate" clade. Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70. Ambiguously aligned regions were trimemed and only genes sampled for at least two-thirds of species were retained. The phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes.

@@

Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatworms. The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL. The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy. With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).

@@

They also investigated saturation, and concluded that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh]. Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.

@@

The final matrix was first analyzed using the Poisson+CAT model. Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is based on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets. Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.

@@

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the GTR+CAT analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 7).

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
  github_document: default
---

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 4). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

Other approaches have been taken to addressing base compositional heterogeneity across taxa. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. However, our own analyses (Supplementary Information section S3, Supplementary Figs. 5-7) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition. Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence. Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many? Moreover, taxa sampling also plays an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.

@@

#### Sensitivity analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_choanozoa (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Chaonimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.

To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above. The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.

@@

We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.

@@

**Extended Data Fig. 3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sample isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis. The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses. (C,D) MDS plots for the amino acid frequencies. Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies. A single MDS analysis was done for all analyses, so that the projections are the same in each plot. For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories. This adds a considerable number of parameters for describing a similar overall space.

@@

**Extended Data Fig. 4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

@@

Note that there was, in fact, an exception to strong support. An analysis of the 40 ribosomal proteins in the matrix recovered Ctenophora-sister with only 69% support. This study did not include the placozoan ingroup *Trichoplax*.

@@

Philippe *et al.* [@Philippe:2009hh] assembled an EST dataset for 55 species with 128 genes to explore the phylogenetic relationship among early diverging animals by adding 9 new species. The data matrix was assembled based on the phylogenetic analysis using Poisson+CAT model, which strongly supported Porifera-sister, and ctenophores were recovered as sister to cnidarians forming the "coelenterate" clade. Gene trees were then constructed, and potentially paralogs were removed by a bootstrap threshold of 70. Ambiguously aligned regions were trimmed and only genes sampled for at least two-thirds of species were retained. The phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes.

@@

Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatworms. The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes which were clustered by MCL. The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy. With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).

@@

They also investigated saturation, and concluded that the Dunn *et al.* [@Dunn:2008ky] data matrix is more saturated than Philippe *et al.* 2009 [Philippe:2009hh]. Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.

@@

The final matrix was first analyzed using the Poisson+CAT model. Different from other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is based on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets. Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removing taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.

@@

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the GTR+CAT analysis of the SR-6 recoded Whelan data. We created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 7).
